# Reproducible Research: Peer Assessment 2
#
## 1. Title: Research Study on Impact of Storms and Severe Weather Events on Public Health and Economy
#
## 2. Synopsis
## In this research study, we have analyzed the impact of storms and other severe weather events on public health and the economy using data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database with events collected from 1950 - 2011. There are two research questions addressed by this study: 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to the population health (i.e., highest mortality rates); 2. Across the United States, which types of events have caused the greatest economic consequences (i.e., highest estimated property damage). The top 3 severe weather events in the U.S. from 1991 - 2011 causing injuries are Tornado, Heavy Wind Storm, and Flood. The top 3 severe weather events in the U.S. from 1991 - 2011 causing fatalities are Heavy Wind Storm, Excessive Heat, and Tornado. The top 3 severe weather events in the U.S. from 1991 - 2011 causing crop damage as a measure of economic consequence are Hot Weather, Heavy Wind Storm, and Flood. The top 3 severe weather events in the U.S. from 1991 - 2011 causing property damage are Heavy Wind Storm, Flood, and Tornado.
#
#

## 3. Data Processing
## This section provides details on the data processing performed on the NOAA dataset.

### 3.1 Basic Global Settings
```{r settings}
echo = TRUE          # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
```

### 3.2 Plotting System and Other Libraries
```{r plotting systems, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)    # for plots 
library(R.utils)    # for bunzip2
library(lubridate)  # for date/time conversions
library(grid)       # for grid graphics package
library(plyr)       # for split-apply paradigm
library(gridExtra)  # for extra functions of grid such as arrange
```


### 3.3 Keeping track of my software environment
```{r environment}
setwd("~/Documents/RepData_PeerAssessment2/")
sessionInfo()
```

### 3.4 Perform download of file using reproducible research best practice of no manual intervention and computer performing all tasks
```{r teach download}
if(!file.exists("~/Downloads/StormData.csv.bz2")){
  download.file(url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile="~/Downloads/StormData.csv.bz2")
}
```

### 3.5 Loading and preprocessing the data as this is a very large dataset, only read it in if it does not exist in environment.
```{r load data, results='markup', warning=TRUE, message=TRUE, cache=TRUE}
if(!file.exists("~/Downloads/StormData.csv")){
    bunzip2(filename="~/Downloads/StormData.csv.bz2", destname="~/Downloads/StormData.csv", overwrite=TRUE, remove=FALSE)
}
if (!"stormData" %in% ls()){
    stormData <- read.csv("~/Downloads/StormData.csv")
}

```
* There are `r dim(stormData)[1]` rows and `r dim(stormData)[2]` columns in the dataset.

### 3.6. Process/transform the data (if necessary) into a format suitable for the analysis. 
#### 3.6.1 Let's remove irrelevant columns and rows. Events in the dataset were collected from 1950 to 2011. 
#### We know more data was collected in the later years looking at the histogram so lets look at 
#### the last 20 years of the data, 1991 to 2011. 
```{r process}
dim(stormData)

smallData <- stormData[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
smallData$YEAR <- as.numeric(year(mdy_hms(smallData[,1])))
hist(smallData$YEAR, main="Histogram of Event Data in US\n from 1950-2011\n", xlab="Year", ylab="Frequency")
smallData <- smallData[smallData$YEAR >= 1991, ]
#
# Let's explore the data first to understand transformations to be made
unique(smallData$YEAR)
unique(smallData$EVTYPE)
unique(smallData$PROPDMGEXP)
unique(smallData$CROPDMGEXP)
```

#### 3.6.2.1 Transformation 1:Set property and crop damage variables to a numeric value based on exponent. Replace invalid expressions with a numeric value of 0.Both PROPDMGEXP and CROPDMGEXP columns contain a  multiplier for each observation for example where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r transform1}
# Transformation function 
transform <- function (fData = smallData, f1, fexp) {
fData[(fData[,fexp] == "+"), f1] <- 0
fData[(fData[,fexp] == "-"), f1] <- 0
fData[(fData[,fexp] == "?"), f1] <- 0

fData[(fData[,fexp] ==  ""),  f1] <- fData[(fData[,fexp] ==  ""),  f1]  * 1e+00
fData[(fData[,fexp] == "0"),  f1] <- fData[(fData[,fexp] == "0"),  f1]  * 1e+00
fData[(fData[,fexp] == "1"),  f1] <- fData[(fData[,fexp] == "1"),  f1]  * 1e+01
fData[(fData[,fexp] == "2"),  f1] <- fData[(fData[,fexp] == "2"),  f1]  * 1e+02
fData[(fData[,fexp] == "3"),  f1] <- fData[(fData[,fexp] == "3"),  f1]  * 1e+03
fData[(fData[,fexp] == "4"),  f1] <- fData[(fData[,fexp] == "4"),  f1]  * 1e+04
fData[(fData[,fexp] == "5"),  f1] <- fData[(fData[,fexp] == "5"),  f1]  * 1e+05
fData[(fData[,fexp] == "6"),  f1] <- fData[(fData[,fexp] == "6"),  f1]  * 1e+06
fData[(fData[,fexp] == "7"),  f1] <- fData[(fData[,fexp] == "7"),  f1]  * 1e+07
fData[(fData[,fexp] == "8"),  f1] <- fData[(fData[,fexp] == "8"),  f1]  * 1e+08
fData[(fData[,fexp] == "9"),  f1] <- fData[(fData[,fexp] == "9"),  f1]  * 1e+09

fData[(toupper(fData[,fexp]) == "H"),  f1] <- fData[(toupper(fData[,fexp]) == "H"),  f1]  * 1e+02
fData[(toupper(fData[,fexp]) == "K"),  f1] <- fData[(toupper(fData[,fexp]) == "K"),  f1]  * 1e+03
fData[(toupper(fData[,fexp]) == "M"),  f1] <- fData[(toupper(fData[,fexp]) == "M"),  f1]  * 1e+06
fData[(toupper(fData[,fexp]) == "B"),  f1] <- fData[(toupper(fData[,fexp]) == "B"),  f1]  * 1e+09
return (fData)
}
dataSet <-transform(smallData, which(colnames(smallData) == "PROPDMG"), which(colnames(smallData) == "PROPDMGEXP"))
dataSet <-transform(smallData, which(colnames(smallData) == "CROPDMG"), which(colnames(smallData) == "CROPDMGEXP"))

```

#### 3.6.2.1 Transformation 2: Group together Event Types (EVTYPE) using grep and ignoring case statements as there are too many  different event types. 

```{r transform2}
# EVTYPE Transformation function 
dataSet$EVTYPE <- as.character(dataSet$EVTYPE)
dataSet$EVTYPE[grepl("tsunami|astronomical|rogue|hazardous|marine|current|tide",
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "TSUNAMI/HIGH CURRENT"
dataSet$EVTYPE[grepl("thunderstorm|lightning|gusty thunderstorm",
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "THUNDERSTORM/LIGHTNING"
dataSet$EVTYPE[grepl("flood", 
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "FLOOD"
dataSet$EVTYPE[grepl("slide|erosion", 
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "LANDSLIDE/EROSION"
dataSet$EVTYPE[grepl("heavy rain | hail|precipitation|burst", 
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "HEAVYRAIN/HAILSTORM"
dataSet$EVTYPE[grepl("freeze|cold|winter weather mix|snow|chill|patchy ice", 
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "COLD WEATHER"
dataSet$EVTYPE[grepl("wind |storm|hurricane|typhoon|tstm wind|high wind", 
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "HEAVY WIND STORM"
dataSet$EVTYPE[grepl("fire|smoke|dense smoke", 
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "FIRE/SMOKE"
dataSet$EVTYPE[grepl("hot|drought|temperature record|record temperature|record high|dry",
                     dataSet$EVTYPE,ignore.case = TRUE)] <- "HOT WEATHER"
dataSet$EVTYPE[grepl("tornado", dataSet$EVTYPE,ignore.case = TRUE)] <- "TORNADO"

```




* After processing, there are `r dim(smallData)[1]` rows and `r dim(smallData)[2]` columns in the dataset.

## 4. Results
### 4.1 Summarize the data across the event types and store the top 5 values
```{r summations}
sumFatalities  <- head(arrange(aggregate(FATALITIES ~ EVTYPE, dataSet, sum), desc(FATALITIES)),n=10)
sumInjuries    <- head(arrange(aggregate(INJURIES ~ EVTYPE,   dataSet, sum), desc(INJURIES)),  n=10)
sumPropDamage  <- head(arrange(aggregate(PROPDMG ~ EVTYPE, dataSet, sum),    desc(PROPDMG)),   n=10)
sumCropDamage  <- head(arrange(aggregate(CROPDMG ~ EVTYPE, dataSet, sum),    desc(CROPDMG)),   n=10)

head(sumFatalities)
head(sumInjuries)
head(sumPropDamage)
head(sumCropDamage)
```

### 4.2 Results of the data analysis to addresses the following questions:

#### 4.2.1 Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health? The results in terms of injuries and fatalities are: 
```{r question1}
gg1 <- ggplot(sumInjuries, aes(EVTYPE, INJURIES)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  scale_y_continuous("Number of Injuries") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Severe Weather Events\n from 1991-2011\n Across the U.S.\n Causing Injuries\n ", x = "Severe Weather Type")

gg2 <- ggplot(sumFatalities, aes(EVTYPE, FATALITIES)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  scale_y_continuous("Number of Fatalities") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Severe Weather Events\n from 1991-2011\n Across the U.S.\n Causing Fatalities\n ", x = "Severe Weather Type")

grid.arrange (gg1, gg2, ncol=2, padding=3)
```


#### The top 3 severe weather events in the U.S. from 1991 - 2011 causing injuries are Tornado, Heavy Wind Storm, and Flood. The top 3 severe weather events in the U.S. from 1991 - 2011 causing fatalities are Heavy Wind Storm, Excessive Heat, and Tornado. 


#### 4.2.2 Across the United States, which types of events have the greatest economic consequences? The results as measured by crop and property damage in terms of US dollars are: 
```{r question2}

gg1 <- ggplot(sumCropDamage, aes(EVTYPE, CROPDMG)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  scale_y_continuous("Crop Damage in US Dollars") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Severe Weather Events\n from 1991-2011\n Across the U.S.\n Impacting Crops\n ", x = "Severe Weather Type")

gg2 <- ggplot(sumPropDamage, aes(EVTYPE, PROPDMG)) + 
  geom_bar(stat = "identity", width = 0.5) + 
  scale_y_continuous("Property Damage in US Dollars") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Severe Weather Events\n from 1991-2011\n Across the U.S.\n Impacting Property\n ", x = "Severe Weather Type")

grid.arrange (gg1, gg2, ncol=2, padding=3)
```

#### The top 3 severe weather events in the U.S. from 1991 - 2011 causing crop damage as a measure of economic consequence are Hot Weather, Heavy Wind Storm, and Flood. The top 3 severe weather events in the U.S. from 1991 - 2011 causing property damage are Heavy Wind Storm, Flood, and Tornado.
