# Reproducible Research: Peer Assessment 2
#
#  Title: Research Study on Impact of Storms and Severe Weather Events on Public Health and Economy
#
## Synopsis
## Storms and other severe weather events may cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. There are two research questions address by this study. First, across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to the population health (i.e., highest mortality rates). Second, across the United States, which types of events have caused the greatest economic consequences (i.e., highest estimated property damage). Some documentation of the data may be found https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf and https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf.



## Basic Global Settings
```{r settings}
echo = TRUE          # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
```

## Plotting System
```{r plotting systems, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
```


### Keeping track of my software environment
```{r environment}
sessionInfo()
```

### Teach computer to do same thing using R
```{r teach download}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "/Users/Terri/Documents/RepData_PeerAssessment2/StormData.csv.bz2")
```

## Loading and preprocessing the data
### 1. Load the data (i.e. read.csv())
```{r load data, results='markup', warning=TRUE, message=TRUE}
if(!file.exists("StormData.csv")){
    unzip("StormData.csv.bz2")
}
stormData <- read.csv("StormData.csv")
```
### 2. Process/transform the data (if necessary) into a format suitable for the analysis
```{r process}
# No processing needed
```


## Your data analysis must address the following questions:

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

## 2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.