# Reproducible Research: Peer Assessment 2

## Title: Research Study on Impact of Storms and Severe Weather Events on Public Health and Economy

## Basic Global Settings
```{r settings}
echo = TRUE          # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
```

## Plotting System
```{r plotting systems, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
```


### Keeping track of my software environment
```{r environment}
sessionInfo()
```

### Teach computer to do same thing using R
```{r teach download}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "/Users/Terri/Documents/RepData_PeerAssessment2/StormData.csv.bz2")
```

## Loading and preprocessing the data
### 1. Load the data (i.e. read.csv())
```{r load data, results='markup', warning=TRUE, message=TRUE}
if(!file.exists("StormData.csv")){
    unzip("StormData.csv.bz2")
}
stormData <- read.csv("StormData.csv")
```
### 2. Process/transform the data (if necessary) into a format suitable for the analysis
```{r process}
# No processing needed
```


## Your data analysis must address the following questions:

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

## 2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.